Simulations of Ballot Polling Risk-Limiting Audits

### 2023-02-07 providence/sections/adversary.tex
.
fix lemma numbering

To illustrate the practical implication of this property, we consider a toy example: an RLA of a two-candidate contest with margin $0.01$ and risk limit $0.1$. 
Suppose we wish to achieve a conditional stopping probability $0.9$ in each round of the audit. For \Providence, we can compute a new round size for each round based on the previous samples. For \Minerva, however, we would have a predetermined round schedule. We use the default \Minerva round schedule of audit software Arlo \cite{arlo} (used by many states performing an RLA), which is is $[x, 2.5x, 6.25x, ...]$; that is, the next marginal round size is $1.5$ times the current one. This multiplier of $1.5$ is known to give, over a wide range of margins, a probability of stopping roughly $0.9$ in the second round if the first round size has probability of stopping $0.9$.

@@

\subsection{Efficiency}
\newtheorem{lem}{Lemma}
\begin{restatable}{lem}{efficiencylemma}
\label{lem:efficiency}
% For any round schedule %$[n_1, \ldots, n_r]$
For any risk-limit $\alpha \in (0, 1)$, for any margin
and for any round schedule $[n_1, \ldots, n_j]$, 
the \Providence RLA is more efficient than EoR \BRAVO.
\end{restatable}
\begin{proof} Appendix~\ref{pf:efficiency_proof}\end{proof}

>>>>

To illustrate the practical implication of this property, we consider a toy example: an RLA of a two-candidate contest with margin $0.01$ and risk limit $0.1$. 
Suppose we wish to achieve a conditional stopping probability $0.9$ in each round of the audit. For \Providence, we can compute a new round size for each round based on the previous samples. For \Minerva, however, we would have a predetermined round schedule. We use the default \Minerva round schedule of audit software Arlo \cite{arlo} (used by many states performing an RLA), which is $[x, 2.5x, 6.25x, ...]$; that is, the next marginal round size is $1.5$ times the current one. This multiplier of $1.5$ is known to give, over a wide range of margins, a probability of stopping roughly $0.9$ in the second round if the first round size has probability of stopping $0.9$.

@@

\subsection{Efficiency}
\begin{lemma}
\label{lem:efficiency}
% For any round schedule %$[n_1, \ldots, n_r]$
For any risk-limit $\alpha \in (0, 1)$, for any margin
and for any round schedule $[n_1, \ldots, n_j]$, 
the \Providence RLA stops before or at the same time as EoR \BRAVO.
\end{lemma}
\begin{proof} Appendix~\ref{sec:proofs}\end{proof}

### 2023-02-07 providence/sections/prov.tex
.

Notice that \Providence requires the computation of $\tau_j$ for $j=1$ and no other values of $j$. The value of $\tau_1$ is simply the ratio of the the tails of the binomial distributions for the two hypotheses and can be fairly efficiently computed. The computation of $\tau_j$ for $j \geq 2$, as required in \Minerva, relies on the convolution of two probability distribution functions and is hence computationally considerably more expensive.

@@

\subsection{Risk-Limiting Property: Proof}
\label{sec:proof}
We now prove that \Providence is risk-limiting using lemmas from basic algebra in Appendix~\ref{sec:proofs}.

@@

For any round $j\ge 2$, by Definition \ref{def:minervatwo}
and Lemma \ref{lemma:minerva2_kmin_exists},
$\mathcal{A}=\text{Correct}$ (the audit stops) if and only if
\begin{equation*}
\begin{aligned}
\omega_{j}(k_{j}, k_{j-1}, p_a, p_0, n_{j}, n_{j-1}, \alpha )\triangleq\\
\sigma(k_{j-1},p_a,p_0,n_{j-1})\cdot \tau_1(k_{j}-k_{j-1},p_a,p_0,n_j-n_{j-1})
\ge \frac{1}{\alpha}.
\end{aligned}
\end{equation*}
%Let $d_j=k_j-k_{j-1}$ be shorthand for the new draw.
%could do this for simpler notation^
By Lemma \ref{lemma:any_ratio_is_sigma_simple}
and Definition \ref{def:minerva_ratio}, this is equivalent to
$$
\frac{\Pr[K_{j-1} = {k_{j-1}} \mid H_a, n_{j-1}]\cdot Pr[K_{j} \ge k_{j} \mid {k_{j-1}}, H_a, n_{j-1}, n_{j}]}{\Pr[K_{j-1} = {k_{j-1}} \mid H_0, n_{j-1}]\cdot Pr[K_{j} \ge k_{j} \mid {k_{j-1}}, H_0, n_{j-1}, n_{j}]}\ge \frac{1}{\alpha}.
$$
By Lemma~\ref{lemma:minerva2_kmin_exists} and Definition~\ref{def:minervatwo},
we see that there exists a $k_{min, j} = k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j}  \leq k_j$ 
%$k_{min, j}\le k_j$ 
for which
$$
\frac{\Pr[K_{j-1} = {k_{j-1}} \mid H_a, n_{j-1}]\cdot Pr[K_{j} \ge k_{j} \mid {k_{j-1}}, H_a, n_{j-1}, n_{j}]}{\Pr[K_{j-1} = {k_{j-1}} \mid H_0, n_{j-1}]\cdot Pr[K_{j} \ge k_{j} \mid {k_{j-1}}, H_0, n_{j-1}, n_{j}]}\ge
$$
$$
\frac{\Pr[K_{j-1} = {k_{j-1}} \mid H_a, n_{j-1}]\cdot Pr[K_{j} \ge k_{min, j} \mid {k_{j-1}}, H_a, n_{j-1}, n_{j}]}{\Pr[K_{j-1} = {k_{j-1}} \mid H_0, n_{j-1}]\cdot Pr[K_{j} \ge k_{min, j} \mid {k_{j-1}}, H_0, n_{j-1}, n_{j}]} \ge 
\frac{1}{\alpha}
$$
%Equivalently,
%$$
%\frac{Pr[K_{j} \ge k_{min, j}\wedge K_{j-1} = k_{j-1} \mid n_{j}^*, k_{j-2}^*, H_a]}{Pr[K_{j}\ge k_{min, j} \wedge K_{j-1} = k_{j-1} \mid n_{j}^*, k_{j-2}^*, H_0]}\ge \frac{1}{\alpha}.
%$$
%Taking the sum over all possible rounds and corresponding preceding values
%of $k$, we get
% Taking the sum over all possible audit histories, we get
The above may be rewritten as
\begin{equation*}
\begin{aligned}
\sum_{{k} = k_{min, j}}^{n_j} Pr[(K_{j} , K_{j-1}) = (k, k_{j-1}) \mid H_0, n_{j-1}, n_{j}] \leq \\
\alpha \sum_{{k} = k_{min, j}}^{n_j} Pr[(K_{j} , K_{j-1}) = (k, k_{j-1}) \mid H_a, n_{j-1}, n_{j}]
\end{aligned}
\end{equation*}
The left hand side above is the probability of stopping in the $j^{th}$ round and $K_{j-1} = k_{j-1}$, given the null hypothesis, which is smaller than $alpha$ times the same probability given the alternate hypothesis. Summing both sides over all values of $K_{j-1} < k_{min, j-1}$ gives us a similar relationship between the probabilities of stopping in round $j$ (given the null and alternate hypotheses respectively). Note here that the relationship holds even if the values of $n_{j}$ depend on $k_{j-1}$. When both sides of the inequality are further summed over all rounds, we get:

>>>>

Notice that \Providence requires the computation of $\tau_j$ for $j=1$ and no other values of $j$. The value of $\tau_1$ is simply the ratio of the the tails of the binomial distributions for the two hypotheses and can be fairly efficiently computed. The computation of $\tau_j$ for $j \geq 2$, as required in \Minerva, relies on the convolution of two probability distribution functions and is hence computationally considerably more expensive. Lesser computational complexity makes audit planning and analysis using simulations as in Section~\ref{sec:workload} more feasible.

@@

\subsection{Risk-Limiting Property: Proof}
\label{sec:proof}
We now prove that \Providence is risk-limiting using lemmas from basic algebra which are given in Appendix~\ref{sec:proofs}.

@@

For any round $j\ge 2$, by Definition \ref{def:minervatwo}
and Lemma \ref{lemma:minerva2_kmin_exists},
$\mathcal{A}=\text{Correct}$ (the audit stops) if and only if
\begin{equation*}
\begin{aligned}
\omega_{j}(k_{j}, k_{j-1}, p_a, p_0, n_{j}, n_{j-1}, \alpha )\triangleq\\
\sigma(k_{j-1},p_a,p_0,n_{j-1})\cdot \tau_1(k_{j}-k_{j-1},p_a,p_0,n_j-n_{j-1})
\ge \frac{1}{\alpha}.
\end{aligned}
\end{equation*}
%Let $d_j=k_j-k_{j-1}$ be shorthand for the new draw.
%could do this for simpler notation^
By Lemma \ref{lemma:any_ratio_is_sigma_simple}
and Definition \ref{def:minerva_ratio}, this is equivalent to
$$
\frac{\Pr[K_{j-1} = {k_{j-1}} \mid H_a, n_{j-1}] Pr[K_{j} \ge k_{j} \mid {k_{j-1}}, H_a, n_{j-1}, n_{j}]}{\Pr[K_{j-1} = {k_{j-1}} \mid H_0, n_{j-1}] Pr[K_{j} \ge k_{j} \mid {k_{j-1}}, H_0, n_{j-1}, n_{j}]}
$$$$\ge \frac{1}{\alpha}.
$$
By Lemma~\ref{lemma:minerva2_kmin_exists} and Definition~\ref{def:minervatwo},
we see that there exists a $k_{min, j} = k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j}  \leq k_j$ 
%$k_{min, j}\le k_j$ 
for which
$$
\frac{\Pr[K_{j-1} = {k_{j-1}} \mid H_a, n_{j-1}] Pr[K_{j} \ge k_{j} \mid {k_{j-1}}, H_a, n_{j-1}, n_{j}]}{\Pr[K_{j-1} = {k_{j-1}} \mid H_0, n_{j-1}] Pr[K_{j} \ge k_{j} \mid {k_{j-1}}, H_0, n_{j-1}, n_{j}]}\ge
$$
$$
\frac{\Pr[K_{j-1} = {k_{j-1}} \mid H_a, n_{j-1}] Pr[K_{j} \ge k_{min, j} \mid {k_{j-1}}, H_a, n_{j-1}, n_{j}]}{\Pr[K_{j-1} = {k_{j-1}} \mid H_0, n_{j-1}] Pr[K_{j} \ge k_{min, j} \mid {k_{j-1}}, H_0, n_{j-1}, n_{j}]} 
$$$$
\ge 
\frac{1}{\alpha}
$$
%Equivalently,
%$$
%\frac{Pr[K_{j} \ge k_{min, j}\wedge K_{j-1} = k_{j-1} \mid n_{j}^*, k_{j-2}^*, H_a]}{Pr[K_{j}\ge k_{min, j} \wedge K_{j-1} = k_{j-1} \mid n_{j}^*, k_{j-2}^*, H_0]}\ge \frac{1}{\alpha}.
%$$
%Taking the sum over all possible rounds and corresponding preceding values
%of $k$, we get
% Taking the sum over all possible audit histories, we get
The above may be rewritten as
\begin{equation*}
\begin{aligned}
\sum_{{k} = k_{min, j}}^{n_j} Pr[(K_{j} , K_{j-1}) = (k, k_{j-1}) \mid H_0, n_{j-1}, n_{j}] \leq \\
\alpha \sum_{{k} = k_{min, j}}^{n_j} Pr[(K_{j} , K_{j-1}) = (k, k_{j-1}) \mid H_a, n_{j-1}, n_{j}]
\end{aligned}
\end{equation*}
The left hand side above is the probability of stopping in the $j^{th}$ round and $K_{j-1} = k_{j-1}$, given the null hypothesis, which is smaller than $alpha$ times the same probability given the alternate hypothesis. Summing both sides over all values of $K_{j-1} < k_{min, j-1}$ gives us a similar relationship between the probabilities of stopping in round $j$ (given the null and alternate hypotheses respectively). Note here that the relationship holds even if the values of $n_{j}$ depend on $k_{j-1}$. When both sides of the inequality are further summed over all rounds, we get:

### 2023-02-07 providence/sections/related.tex
shortening
.

%If testing the \BRAVO stopping condition after each individual ballot
% is drawn (a \B \BRAVO audit), 
It is easy to see that the ratio $\sigma$ is the 
likelihood ratio:
$$
\frac{Pr[K=k|H_a,n]}{Pr[K=k|H_0,n]}= \frac{\binom{n}{k}p_a^{k} (1-p_a)^{n-k}}{\binom{n}{k}(\frac{1}{2})^n} =\sigma(k, p_a, \frac{1}{2}, n)
$$

@@

It now becomes useful to have shorthand for a sequence of cumulative round sizes and the corresponding sequence
of cumulative winner ballot tallies.
We use:
$$\bm{k_j}\triangleq(k_1,k_2,\ldots,k_j)$$
$$\bm{n_j}\triangleq(n_1,n_2,\ldots,n_j)$$

Where \BRAVO uses the ratio of the values of the probability distribution functions, \Minerva uses the ratio of their \emph{tails}.

>>>>

%If testing the \BRAVO stopping condition after each individual ballot
% is drawn (a \B \BRAVO audit), 
If $K$ is the random variable indicating the number of ballots in the sample that contain a vote for the reported winner, it is easy to see that the ratio $\sigma$ is the likelihood ratio:
$$
\frac{Pr[K=k|H_a,n]}{Pr[K=k|H_0,n]}= \frac{\binom{n}{k}p_a^{k} (1-p_a)^{n-k}}{\binom{n}{k}(\frac{1}{2})^n} =\sigma(k, p_a, \frac{1}{2}, n)
$$

@@

Where \BRAVO uses the ratio of the values of the probability distribution functions, \Minerva uses the ratio of their \emph{tails}. Now it becomes useful to have shorthand for a sequence of cumulative round sizes and the corresponding sequence
of cumulative winner ballot tallies.
We use:
$$\bm{n_j}\triangleq(n_1,n_2,\ldots,n_j) \quad\text{and}\quad \bm{k_j}\triangleq(k_1,k_2,\ldots,k_j)$$
Also, let $K_j$ be the random variable indicating the cumulative number of ballots in the sample after the $j$th round is drawn.

### 2023-02-07 providence/sections/workload.tex
.

Figure~\ref{fig:avg_bals_ratio} provides the same number as a fraction of the \Providence values.
It is straightforward to show that \Providence and both forms of \BRAVO collapse to the same test when each round corresponds to a single ballot. Figures~\ref{fig:avg_bals} and \ref{fig:avg_bals_ratio} show that for larger stopping probabilities $p$ (i.e. larger rounds), \Providence requires fewer ballots on average. In particular, the savings of \Providence become larger as $p$ increases; for $p=.95$, EoR \BRAVO and SO \BRAVO require more than $2$ and $1.4$ times as many ballots as \Providence respectively. 
\begin{figure}[h!]
\includegraphics[width=.5\textwidth]{avg_bals_ratio.png}
\caption{The total number of ballots sampled on average, as a fraction of those sampled by \Providence, as a function of $p$, the conditional stopping probability used to select each round size, for ballot polling audits of the 2020 US Presidential election in the state of Virginia.}
\label{fig:avg_bals_ratio}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=.5\textwidth]{with_round_workload.png}
\caption{For workload parameters $w_b=1$ and $w_r=1000$, this plot shows the expected workload for various round schedule parameters $p$. Expected workload is found using Equation~\ref{eq:round_workload} and the average number of ballots and rounds in our simulations as the expected number of ballots and rounds.}
\label{fig:with_round_workload}
\end{figure}

@@

\subsubsection{Precinct overhead.} For a more complete model, we can also introduce container-level workload. If. around requires multiple ballots from a single container, the container need only be unsealed once. Based on a Rhode Island pilot RLA report\cite{RI-report}, this may mean that a ballot from a new container requires roughly twice the time as a ballot from an already-opened container. Typically available election results give per-precinct granularity of vote tallies, rather than individual container information. In Virginia, however, most precincts have a single ballot scanner whose one box has sufficient capacity for all the ballots cast in that precinct anyways, and so we model the per-container workload as a per-precinct workload, $w_p$. In this model, the workload estimate incurs an additional workload of $w_p$ every time a precinct is sampled from for the first time in a round. That is, let $E_{pi}$ be the expected number of distinct precincts sampled from in round $i$, and let $E_p=\sum_i E_{pi}$. Then the new model is
\begin{equation}
W(E_b, E_r, E_p) = E_b w_b + E_r w_r + E_p w_p + C
\label{eq:round_and_precinct_workload}
\end{equation}

@@

\subsection{Real time}
Given tight certification deadlines\footnote{Virginia recently passed legislation requiring pre-certification RLAs.}, the total real time to conduct the RLA is also an important factor to consider when planning audits.
Because each county can sample ballots for the same round concurrently, the total real time for a round depends only on the slowest county. 
In Virginia, Fairfax County typically has the most votes cast by a significant difference; in the contest we consider, Fairfax County had ~551 thousand votes cast, more than double the ~203 thousand of second-highest Virginia Beach City.
Consequently, we model the expected total real time $T$ of an audit using just the largest county, and we define analagous variables for the expected values in just the largest county.
For the largest county, let the expected total ballots sampled be $\bar E_b$, the expected number of rounds $\bar E_r$, and the expected number of distinct precinct samples summed over all rounds be $\bar E_c$.
Similarly, we use real time per-ballot, per-round, and per-precinct workload variables, $t_b$, $t_r$, and $t_p$. So the real time of the audit is estimated by
\begin{equation}
T(\bar E_b, \bar E_r, \bar E_p ) = \bar E_b t_b + \bar E_r t_r + \bar E_p t_p + C
\label{eq:real_time}
\end{equation}

@@

We observe a similar behavior in our simulations of audits on the contest from the pilot audit. Figure~\ref{fig:pilot_misleading} shows the proportion of the pilot simulations which contained a \emph{misleading sample} in any round. Despite the large difference in margin ($\sim 5\%$ in Virginia and $\sim 25\%$ in the pilot) we still observe that a \emph{misleading limit} of $0.01$ is first achieved at roughly $p=0.8$ and $0.001$ at $p=0.95$.

>>>>

Figure~\ref{fig:avg_bals_ratio} provides the same number as a fraction of the \Providence values.
It is straightforward to show that \Providence and both forms of \BRAVO collapse to the same test when each round corresponds to a single ballot. Figures~\ref{fig:avg_bals} and \ref{fig:avg_bals_ratio} show that for larger stopping probabilities $p$ (i.e. larger rounds), \Providence requires fewer ballots on average. In particular, the savings of \Providence become larger as $p$ increases; for $p=0.95$, EoR \BRAVO and SO \BRAVO require more than $2$ and $1.4$ times as many ballots as \Providence respectively. 
\begin{figure}[h!]
\includegraphics[width=.5\textwidth]{avg_bals_ratio.png}
\caption{The total number of ballots sampled on average, as a fraction of those sampled by \Providence, as a function of $p$, the conditional stopping probability used to select each round size, for ballot polling audits of the 2020 US Presidential election in the state of Virginia.}
\label{fig:avg_bals_ratio}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=.5\textwidth]{with_round_workload.png}
\caption{For workload parameters $w_b=1$ and $w_r=1000$, this plot shows the expected workload for various values of $p$. Expected workload is found using Equation~\ref{eq:round_workload} and the average number of ballots and rounds in our simulations as the expected number of ballots and rounds.}
\label{fig:with_round_workload}
\end{figure}

@@

\subsubsection{Precinct overhead.} For a more complete model, we can also introduce container-level workload. If a round requires multiple ballots from a single container, the container need only be unsealed once. Based on a Rhode Island pilot RLA report\cite{RI-report}, this may mean that a ballot from a new container requires roughly twice the time as a ballot from an already-opened container. Typically available election results give per-precinct granularity of vote tallies, rather than individual container information. In Virginia, however, most precincts have a single ballot scanner whose one box has sufficient capacity for all the ballots cast in that precinct anyways, and so we model the per-container workload as a per-precinct workload, $w_p$. In this model, the workload estimate incurs an additional workload of $w_p$ every time a precinct is sampled from for the first time in a round. That is, let $E_{pi}$ be the expected number of distinct precincts sampled from in round $i$, and let $E_p=\sum_i E_{pi}$. Then the new model is
\begin{equation}
W(E_b, E_r, E_p) = E_b w_b + E_r w_r + E_p w_p + C
\label{eq:round_and_precinct_workload}
\end{equation}

@@

\subsection{Real time}
Given tight certification deadlines\footnote{Virginia recently passed legislation requiring pre-certification RLAs.}, the total real time to conduct the RLA is also an important factor to consider when planning audits.
Because each county can sample ballots for the same round concurrently, the total real time for a round depends only on the slowest county. 
In Virginia, Fairfax County typically has the most votes cast by a significant difference; in the contest we consider, Fairfax County had ~551 thousand votes cast, more than double the ~203 thousand of second-highest Virginia Beach City.
Consequently, we model the expected total real time $T$ of an audit using just the largest county, and we define analagous variables for the expected values in just the largest county.
For the largest county, let the expected total ballots sampled be $\bar E_b$, the expected number of rounds $\bar E_r$, and the expected number of distinct precinct samples summed over all rounds be $\bar E_p$.
Similarly, we use real time per-ballot, per-round, and per-precinct workload variables, $t_b$, $t_r$, and $t_p$. So the real time of the audit is estimated by
\begin{equation}
T(\bar E_b, \bar E_r, \bar E_p ) = \bar E_b t_b + \bar E_r t_r + \bar E_p t_p + C
\label{eq:real_time}
\end{equation}

@@

We observe a similar behavior in our simulations of audits on the contest from the pilot audit. Figure~\ref{fig:pilot_misleading} shows the proportion of the pilot simulations which contained a \emph{misleading sample} in any round. Despite the large difference in margin ($\sim 0.05$ in Virginia and $\sim 0.25$ in the pilot) we still observe that a \emph{misleading limit} of $0.01$ is first achieved at roughly $p=0.8$ and $0.001$ at $p=0.95$.
